### 2025-04-09 chapters/2-particles.tex
T-1 day to initial submission, nearly complete draft up to citations and small comments.

The main triumph of this chapter is the introduction of \vocab{partons}:
%
the main actors of modern particle physics whose main role is to compose all other matter, and can be thought of as the fundamental moving pieces in a particle collision experiment.
%
As we introduce the basics of the \glsfirst{sm} and \glsfirst{qcd}, we will discuss what we know about particles and partons, and how we know it.
%
We focus on quarks and gluons, the fundamental particles described by \gls{qcd} (our theory of the strong force) which are the central focus of this thesis.
%
Throughout this thesis, we will also run into the photon, the \(W\)-boson, and even \textit{composite} particles (such as the proton and neutron) that are themselves are composed of quarks, gluons, photons, and more.
%
We will make some quantitative predictions which accentuate the profound strangeness of quantum field theory and foretell the emergence of high-energy jets -- proxies for partonic degrees of freedom that we discuss in the next chapter, and which themselves contain many, many particles.

@@

Experimental and theoretical studies of scattering are usually phrased in terms of \glspl{xsec}, differential cross sections, or probability distributions associated with scattering outcomes.
%
The \vocab{cross section} associated with a certain scattering outcome, denoted by \(\sigma\), characterizes how strong the interactions producing the outcome are.
%
The larger the cross section, the stronger the interactions.
%
\vocab{Differential cross sections} and \vocab{probability distributions} is a more detailed function which characterizes how much of the cross section is associated with particular regions of phase space for the scattering outcome.

@@

Quantum field theory (\gls{qft}) is our simplest framework for relativistic quantum mechanics, and provides elegant solutions to a huge number of conceptual difficulties.%
\footnote{
    See, for example \sam{Daniel's notes} \sam{Neumeier's blog}.
}
%
\Gls{qft} was developed first as a tool to describe the fundamental constituents of our universe -- a story that has culminated in the development of \gls{qcd} and the \gls{sm}.%
\footnote{
    And, depending on your company, string theory.
}
%
Today, however, \gls{qft} is a powerful framework for quantum mechanics even in a wide variety of observed phenomena, including non-relativistic phenomena ranging from nuclear transitions to superfluidity and superconductivity.

Our focus in this thesis, however, remains on the smallest building blocks of the \gls{sm} and \gls{qcd}, which we introduce in \Sec{sm-qcd}, and therefore on the framework of \gls{scattering} we use to understand them, which we introduce in \Sec{scattering}.
%
In order to apply standard \gls{qft} techniques to \gls{qcd} scattering, we will introduce the principle of factorization in \Sec{universal-features}.
%
Finally, with the goal of designing well-behaved theoretical probes of the physics of \gls{qcd} \gls{scattering}, we discuss \glslink{irc-safety}{infra-red and collinear (IRC) safety} in \Sec{irc-safety}.
%
The discussion of this section will leave us ready for whatever may come in the concrete calculations that follow in the remainder of this thesis.

@@

\epigraph{What could be more natural than to unite these spin-one bosons into a multiplet of gauge fields?}{Steven Weinberg, \textit{A Model of Leptons} \cite{}, 1967}

@@

\noindent
Nonetheless, we will take this opportunity to discuss the more general framework of gauge theory and the \gls{sm}.
%
Those who are champing at the bit for \gls{qcd}, and are familiar with the \gls{qcd} Feynman rules, may want to skip immediately to later sections, such as \Sec{scattering} for a discussion of scattering, \Sec{observables} for a discussion of collider observables, or even further to \Sec{universal-features} for a discussion of the universal features of partonic scattering processes.

The \gls{sm} is ultimately an elegant and extremely non-trivial extension of the theory of electromagnetism.
%
There is an infinity of concepts within the \gls{sm} that we will not cover -- we will ignore higher dimension operators, the reasoning behind the classical and quantum geometric structures related to Yang-Mills theory, explicit loop-level computations of quantum effects, and an enormous amount of beautiful phenomenology and formal theory, for a start.
%
Instead, we focus specifically on the physics that will take us to an understanding of jets.
%
We will take a shortcut to the \gls{sm} through the theory of electromagnetism, before focusing exclusively on \gls{qcd}.

\remark{}{
    A more in-depth exposition of \gls{qft} would require a great deal of effort, and distract from the physics of quarks, gluons, and jets that we wish to explore in this thesis.
    %
    Nonetheless, I mention that some of my favorite introductory resources on \gls{qft} are Mark Srednicki's \underline{Quantum Field Theory} \cite{} and Daniel Harlow's \underline{Notes on Quantum Field Theory} \cite{}.
    %
    The latter in particular gives a sense of the full, deep flavor of \gls{qft} from algebraic, calculational, and intuitive perspectives through rigorous and detailed discussion.
}

The theory of electromagnetism describes the physics of electric and magnetic fields is often described by an energy functional/Hamiltonian dependent on the electric and magnetic fields.
%
However, it is often far easier to use \textit{gauge fields} to describe the physics of the electric and magnetic fields.
%
The use of gauge fields make manifest the presence of the symmetries of special relativity -- the Lorentz symmetries combining rotations and boosts -- in the theory of electromagnetism.
%
The gauge fields of electromagnetism are the \textit{electric potential} \(\phi\) and the \textit{vector potential} \(\vec{A}\), which may be grouped into a Lorentz four-vector
\begin{align}
    A_\mu &\reppedby \le(-\phi, \, \vec{A}\ri)
    \,,
\end{align}
which encodes the electric and magnetic fields in the \vocab{field strength tensor},
\begin{align}
    F_{\mu\nu} &= \partial_\mu A_\nu - \partial_\nu A_\mu
    \,.
\end{align}
%
The electric and magnetic fields take the form
\begin{align}
    E^i &= F^{0i} = \partial_t A^i - \partial^i \phi
    \\
    B^i
    &=
    \half \epsilon^{ijk} F^{jk}
    =
    \le(\nabla \cross A\ri)^i
    \,,
\end{align}
\sam{check}
%
The principles of special relativity, and hence the use of gauge fields, will be especially important for us as we study high-energy particle collisions with outgoing particles moving near the speed of light.
%
Gauge fields also allow a more easy generalization of quantum electrodynamics (QED), with electric and magnetic fields, to \gls{qcd}, with the creatively named chromo-electric and chromo-magnetic fields.
%
In particular, under the U(1) \vocab{gauge transformation}
\begin{align}
    \le(\partial_\mu - i e A_\mu\ri)^\prime
    :=
    e^{-i e \alpha(x)}
    \le(\partial_\mu - i e A_\mu\ri)
    e^{i e \alpha(x)}
    \,,
\end{align}
where \(\alpha(x)\) is arbitrary, the gauge field is transformed into
\begin{align}
    A^\prime_{\mu}
    =
    A_\mu
    -
    \partial_\mu \alpha
    \,.
\end{align}
%
This transformation is called a U(1) gauge transformation because the factor \(U := U(x) := e^{-i e \alpha(x)}\) is a \(1 \times 1\) \textit{unitary matrix}, with \(U^\dagger U = 1\), and U(1) is the group whose multiplication structure is isomorphic to the set of 1\(\times\)1 unitary matrices.
%
We sometimes say that \(U = e^{-i e \alpha(x)}\) corresponds to a representation of U(1) at every point \(x\), called the U(1) \textit{gauge group}, with charge \(e\).
%
The object \(D_\mu := \partial_\mu - i e A_\mu\) is called the \textit{covariant derivative} (in the representation of U(1) with charge \(e\)).

@@

Including a massless electron \(e\) and the positron \(\overline{e}\) -- Weyl fermion fields whose properties are discussed in standard texts on quantum field theory (see, e.g. Mark Srednicki's \underline{Quantum Field Theory}, Part II \cite{}) -- yields the Lagrangian for massless \textit{quantum electrodynamics} (QED):
\begin{align}
    \mathcal{L}_\text{QED}
    =
    -\frac{1}{4}F_{\mu\nu} F^{\mu\nu}
    +
    i \overline{e} \, \overline{\sigma}^\mu D_\mu \, e
    =
    -\frac{1}{4}F_{\mu\nu} F^{\mu\nu}
    +
    i \bar{\Psi} \slashed{D} \Psi
    \,,
\end{align}
where \(\Psi = \le(e, \overline{e}\ri)\) is a Dirac spinor that groups together the left- and right-handed Weyl spinors \(e\) and \(\overline{e}\).
%
\(\overline{\sigma}^\mu \reppedby \le(I, \vec{\sigma}\ri)\) is a four-vector whose components are each 2\(\times\)2 matrices;
%
it is an \vocab{invariant symbol}%
\footnote{
    Also known as a Clebsch-Gordan coefficient of the Lorentz group;
    %
    see \href{https://web.physics.ucsb.edu/~mark/ms-qft-DRAFT.pdf\#page=415}{Mark Srednicki's \underline{Quantum Field Theory}, Chapters 34, 35, and 70.} \cite{}.
}
%
of the Lorentz group.
%
The Lagrangian of massless QED contains a free photon (gauge field), encoded by the piece \(\mathcal{L} \supset -F_{\mu\nu}F^{\mu\nu}/4\), a free fermion, encoded by \(\mathcal{L} \supset i\overline{e} \overline{\sigma}^\mu \partial_\mu e = i \bar{\Psi} \slashed{\partial} \Psi\), and an interaction \(\mathcal{L} \supset A^\mu J_\mu = i\overline{e} \overline{\sigma}^\mu A_\mu e = i \bar{\Psi} \slashed{A} \Psi\) between the photon and the electromagnetic current generated by the presence of the electron and positron.

The generalization of electromagnetism that appears in the standard model is known as \textit{Yang-Mills theory}.
%
In Yang-Mills theory, the group U(1) which encodes the gauge transformation properties of \(A_\mu\) is upgraded to a more general group (in particular, a Lie group).
%
The most common example -- and the example relevant for \gls{qcd} and the \gls{sm} -- is the group SU(N), isomorphic to the set of \(N\times N\) unitary matrices with determinant 1.

\remark{}{
    The classical properties of Yang-Mills theory, even with matter (electrons in QED, quarks in \gls{qcd}, and even the Higgs boson in the \gls{sm}), can be explained elegantly in terms of geometric concepts (connections, principal bundles, horizontal subspaces, spinor bundles, etc.).
    %
    Such an explanation is far outside the scope of our discussion, but some relevant mathematical resources include \Reffs{}.
    % Hamilton, spinorial chessboard, gauge theory book
}

@@

The covariant derivative in the representation \(R\) also becomes a matrix, and takes the form
\begin{align}
    D_{(R)\,\mu}
    :=
    \partial_\mu
    -
    i g T^a_{(R)} A^a_\mu
    \equiv
    \partial_\mu
    -
    i g A_{(R)\,\mu}
    \,,
\end{align}
\sam{fix}
where the gauge transformation now takes the form
\begin{align}
    D^\prime_{(R)\,\mu}
    :=
    U_{(R)}
    D_{(R)\,\mu}
    U^\dagger_{(R)}
    \,,
\end{align}
and \(U_{(R)} := U_{(R)}(x)\) is dependent on the space-time position at which it is evaluated.
%
In terms of the gauge field \(A^a_\mu\), we have
\begin{align}
    T^a_{(R)} A^{\prime\,a}_\mu
    =
    A_\mu^b
    U_{(R)} T^b_{(R)} U^\dagger_{(R)}
    +
    \frac{i}{g} U_{(R)} \partial_\mu U^\dagger_{(R)}
    \,,
\end{align}
and, when the \(\alpha^a(x)\) are small enough to discard terms of \(\mathcal{O}(\alpha^2)\),
\begin{align}
    A^{\prime\,a}_\mu
    =
    A^{a}_\mu
    -
    \partial_\mu \alpha^a
    +
    g f^{abc} A^b_\mu \alpha^c
    \,.
\end{align}
\sam{check}
% \sam{plus sign in final term on pg 5 of Nate's 229B Lecture 12}

The field strength tensor is
\begin{align}
    \label{eq:non-abelian-field-strength}
    F_{(R)\,\mu\nu}
    \equiv
    T^a_{(R)}
    F\indices{^a_{\mu\nu}}
    :=
    \frac{i}{g}
    \comm{D_{(R)\,\mu}}{D_{(R)\,\nu}}
    =
    T^a_{(R)}
    \le(
        \partial_\mu A^a_\nu - \partial_\nu A^a_\mu
        + g f^{abc} A^b_\mu A^c_\nu
    \ri)
    \,.
\end{align}
The additional non-linearity of \(F\indices{^a_{\mu\nu}}\) may appear complicated or even innocuous in \Eq{non-abelian-field-strength}, but it has incredibly deep implications for the existence of \gls{qcd} -- it is the reason \gls{qcd} overcomes the weaknesses of the quantum field theories that precede it, and leads to the asymptotic freedom of quarks and gluons discussed in \Sec{phd}.

Matter also matters.
%
Quarks in particular will play a large role in our discussion.
%
A \textit{matter field} \(\varphi\) -- e.g. a scalar or fermion field -- \vocab{transforms under the representation \(R\)} (of the gauge group) if
\begin{align}
    \varphi^\prime(x) = U_{(R)}(x) \, \varphi(x)
    \,.
\end{align}
%
A gauge theory with fermions \(\{\psi_{i}\}\) transforming in the representations \(R_i\) and scalars \(\{\phi_k\}\) transforming in the representations \(R_k\) in \(d=3+1\) dimensions and ignoring higher dimension operators can be described in terms of the \vocab{Yang-Mills-matter Lagrangian}%
\footnote{
    Non-Lagrangian methods for understanding \gls{qft}, such as bootstrap methods (e.g. the \textit{amplitudes program} of \gls{qft}) and the operator product expansion, are also available.
    %
    They can provide more rigorous descriptions of quantum dynamics than we pursue in this thesis, but are often more technically difficult as well.
}.
%
With massless fermions (guaranteed in the \gls{sm} due to the representations of the fermionic \gls{sm} matter content), it takes the form
\begin{subequations}
    \label{eq:yang-mills-matter-lagrangian}
\begin{align}
    \mathcal{L}
    &=
    \mathcal{L}_\text{YM}
    +
    \mathcal{L}_\text{fermion}
    +
    \mathcal{L}_\text{scalar}
    +
    \mathcal{L}_\text{Yukawa}
    \,,
    \\
    \mathcal{L}_\text{YM}
    &=
    - \frac{1}{4} F\indices{^a_{\mu\nu}} F\indices{^{a\,\mu\nu}}
    =
    - \frac{1}{2} \Tr F_{\mu\nu} F^{\mu\nu}
    \,,
    \\
    \mathcal{L}_\text{fermion}
    &=
    i \sum_i \psi_i^\dagger \overline{\sigma}^\mu D_{(R_i)\,\mu} \psi_i
    \\
    \mathcal{L}_\text{scalar}
    &=
    \sum_k \le(D^\mu \phi_k\ri)^\dagger D_\mu \phi_k
    -
    \sum_{k,\ell} \! {}^\prime
    \,\,
    m_{k\ell}^2
    \,\,
    \phi_k \, \phi_\ell
    \,\,
    S_{R_k R_\ell}
    +
    \text{h.c}
    \\
    \notag
    &\qquad\qquad
    +
    \sum_{k,\ell,m,n} \!\!\!\! {}^\prime
    \,\,
    \lambda_{k\ell m n}
    \,\,
    \phi_k \, \phi_\ell \, \phi_m \, \phi_n
    \,\,
    S_{R_k R_\ell R_m R_n}
    +
    \text{h.c.}
    \\
    \mathcal{L}_\text{Yukawa}
    &=
    \sum_{ij}
    \sum_{k} \! {}^\prime
    \,\,
    y_{ijk}
    \,\,
    \psi_i \,\psi_j \,\, \phi_k
    \,\,
    S_{R_i R_j R_k}
    +
    \text{h.c.}
    \,,
\end{align}
\end{subequations}
where the \(S\) are invariant symbols of the gauge group which project onto singlet representations and guarantee the invariance of the Lagrangian under gauge transformations, the \(\sum'\) indicates a sum on both the fields \(\phi_k\) and the complex conjugates \(\phi_k^\dagger\), and ``h.c.'' means hermitian conjugate.

@@

\remark{}{
    \sam{Additional conditions, anomaly cancellation}.
    %
    Nonetheless, the \gls{sm} is still a consistent theory when the top quark is integrated out, though this effective theory has an anomaly.
    %
    For an understanding of the properties of effective field theories which appear to carry anomalies, see the incredibly lucid account of \Reff{}.
}

>>>>

The main triumph of this chapter is the introduction of \vocab{partons}:
%
the main actors of modern particle physics whose main role is to compose all other matter, and can be thought of as the fundamental moving pieces in a particle collision experiment.
%
As we introduce the basics of the \glsfirst{sm} and \glsfirst{qcd}, we will discuss what we know about particles and partons, and how we know it.
%
We focus on quarks and gluons, the fundamental particles described by \gls{qcd} (our theory of the strong force) which are the central focus of this thesis.
%
Throughout this thesis, we will also run into the photon, the \(W\)-boson, and even \textit{composite} particles (such as the proton and neutron) that themselves are composed of quarks, gluons, photons, and more.
%
We will make some quantitative predictions which accentuate the profound strangeness of quantum field theory and foretell the emergence of high-energy jets -- proxies for partonic degrees of freedom that we discuss in the next chapter, and which themselves contain many, many particles.

@@

Experimental and theoretical studies of scattering are usually phrased in terms of \glspl{xsec}, differential cross sections, or probability distributions associated with scattering outcomes.
%
The \vocab{cross section} associated with a certain scattering outcome, denoted by \(\sigma\), characterizes how strong the interactions producing the outcome are.
%
The larger the cross section, the stronger the interactions.
%
\vocab{Differential cross sections} and \vocab{probability distributions} are more detailed functions which characterizes how much of the cross section is associated with particular regions of phase space for the scattering outcome.

@@

Quantum field theory (\gls{qft}) is our simplest framework for relativistic quantum mechanics, and provides elegant solutions to a huge number of conceptual difficulties.%
\footnote{
See, for example, the early chapters of Daniel Harlow's  \textit{Notes on Quantum Field Theory} \cite{}, as well as \href{https://arnold-neumaier.at/physfaq/topics/multi.html}{the blog post of Arnold Neumeier on multiparticle relativistic quantum mechanics, \texttt{available at this https url}}.
}
%
\Gls{qft} was developed first as a tool to describe the fundamental constituents of our universe -- a story that has culminated in the development of \gls{qcd} and the \gls{sm}.%
\footnote{
    And, depending on your company, string theory.
}
%
Today, however, \gls{qft} is a powerful framework for quantum mechanics even in a wide variety of observed phenomena, including non-relativistic phenomena ranging from nuclear transitions to superfluidity and superconductivity.

Our focus in this thesis, however, remains on the smallest building blocks of the \gls{sm} and \gls{qcd}, which we introduce in \Sec{sm-qcd}, and therefore on the framework of \gls{scattering} we use to understand them, which we introduce in \Sec{scattering}.
%
In \Sec{observables}, we introduce \glspl{observable} as theoretical probes of the physics of scattering.
%
In \Sec{irc-safety}, we introduce \glslink{irc-safety}{infra-red and collinear (IRC) safety} as a property of the (perturbatively) well-behaved \glspl{observable} we use in this thesis..
%
\Sec{phd}, which may be skipped on a first reading, discusses why we can trust perturbative calculations involving quarks and gluons, given that the \gls{qcd} matter of our universe is bound up into hadrons.
%
The discussion of this section will leave us ready for whatever may come in the concrete calculations that follow in the remainder of this thesis.

@@

\epigraph{What could be more natural than to unite these spin-one bosons into a multiplet of gauge fields?}{Steven Weinberg, \textit{A Model of Leptons} \cite{Weinberg:1967tq}, 1967}

@@

Readers who are interested in pursuing the theory of \gls{qcd} directly may wish to proceed directly to the \gls{qcd} Feynman rules in \Eqs{gluon_main_frules}{quark_main_frules}, the discussion of particle collisions and collider observables in \Secs{scattering}{observables}, the discussion of partonic scattering in \Sec{universal-features}, or even our discussion of jets in \Chap{jets}.
%
Nonetheless, we will take this opportunity to discuss the more general framework of gauge theory and the \gls{sm}.

The \gls{sm} is ultimately an elegant and extremely non-trivial extension of the theory of electromagnetism.
%
There is an infinity of concepts within the \gls{sm} that we will not cover -- we will not perform explicit loop-level calculations of quantum effects, and we will ignore higher dimension operators, the hierarchy problem (see especially \Reff{Koren:2020pio}), discrete  symmetries (see \Reff{Koren:2022bam}, also with a beautiful explanation of proton stability in the \gls{sm}), and an enormous amount of beautiful phenomenology and formal theory, for a start.
%
The \gls{sm} has a rich formal structure that we will ignore as well, from the classical and quantum geometric structures of Yang-Mills theory to the modern developments of non-invertible symmetries (see e.g. \Reffs{Cordova:2022fhg,Cordova:2022qtz}).
%
Instead, we focus specifically on the physics that will take us to an understanding of jets.
%
We will take a shortcut to the \gls{sm} through the theory of electromagnetism, before focusing exclusively on \gls{qcd}.

\remark{}{
    A more in-depth exposition of \gls{qft} would require a great deal of effort, and distract from the physics of quarks, gluons, and jets that we wish to explore in this thesis.
    %
    Nonetheless, I mention that some of my favorite introductory resources on \gls{qft} are Mark Srednicki's \underline{Quantum Field Theory} \cite{Srednicki:2007qs} and Daniel Harlow's \textit{Notes on Quantum Field Theory} \cite{}.
    %
    The latter in particular gives a sense of the full, deep flavor of \gls{qft} from algebraic, calculational, and intuitive perspectives through rigorous and detailed discussion.
}

The theory of electromagnetism describes the physics of electric and magnetic fields is often described by an energy functional/Hamiltonian dependent on the electric and magnetic fields.
%
However, it is often far easier to use \textit{gauge fields} to describe the physics of the electric and magnetic fields.
%
The use of gauge fields make manifest the presence of the symmetries of special relativity -- the Lorentz symmetries combining rotations and boosts -- in the theory of electromagnetism.
%
The gauge fields of electromagnetism are the \textit{electric potential} \(\phi\) and the \textit{vector potential} \(\vec{A}\), which may be grouped into a Lorentz four-vector
\begin{align}
    A^\mu &\reppedby \le(\phi, \, \vec{A}\ri)
    \,,
\end{align}
which encodes the electric and magnetic fields in the \vocab{field strength tensor},
\begin{align}
    F_{\mu\nu} &= \partial_\mu A_\nu - \partial_\nu A_\mu
    \,.
\end{align}
%
The electric and magnetic fields take the form%
\footnote{
    Note the unfortunate use of the \((+,-,-,-)\) metric.
}
\begin{align}
    E^i &= F_{0i} = -\partial_t A^i - \partial^i \phi
    \\
    B^i
    &=
    \half \epsilon^{ijk} F_{jk}
    =
    \le(\nabla \cross A\ri)^i
    \,,
\end{align}
%
The principles of special relativity, and hence the use of gauge fields, will be especially important for us as we study high-energy particle collisions with outgoing particles moving near the speed of light.
%
Gauge fields also allow a more easy generalization of quantum electrodynamics (QED), with electric and magnetic fields, to \gls{qcd}, with the creatively named chromo-electric and chromo-magnetic fields.
%
In particular, under the U(1) \vocab{gauge transformation}
\begin{align}
    \le(\partial_\mu - i e A_\mu\ri)^\prime
    :=
    e^{-i e \alpha(x)}
    \le(\partial_\mu - i e A_\mu\ri)
    e^{i e \alpha(x)}
    \,,
\end{align}
where \(\alpha(x)\) is arbitrary, the gauge field is transformed into
\begin{align}
    A^\prime_{\mu}
    =
    A_\mu
    -
    \partial_\mu \alpha
    \,.
\end{align}
%
This transformation is called a U(1) gauge transformation because the factor \(U := U(x) := e^{-i e \alpha(x)}\) is a \(1 \times 1\) \textit{unitary matrix}, with \(U^\dagger U = 1\), and U(1) is the group whose multiplication structure is isomorphic to the set of 1\(\times\)1 unitary matrices.
%
We sometimes say that \(U = e^{-i e \alpha(x)}\) corresponds to a representation of U(1) at every point \(x\), called the U(1) \textit{gauge group}, with charge \(e\).
%
The object \(D_\mu := \partial_\mu - i e A_\mu\) is called the \textit{covariant derivative} (in the representation of U(1) with charge \(e\)).

@@

Including a massless electron \(e\) and the positron \(\overline{e}\) -- Weyl fermion fields whose properties are discussed in standard texts on quantum field theory (see, e.g. Mark Srednicki's \underline{Quantum Field Theory}, Part II \cite{Srednicki:2007qs}) -- yields the Lagrangian for massless \textit{quantum electrodynamics} (QED):
\begin{align}
    \mathcal{L}_\text{QED}
    =
    -\frac{1}{4}F_{\mu\nu} F^{\mu\nu}
    +
    i \overline{e} \, \overline{\sigma}^\mu D_\mu \, e
    =
    -\frac{1}{4}F_{\mu\nu} F^{\mu\nu}
    +
    i \bar{\Psi} \slashed{D} \Psi
    \,,
\end{align}
where \(\Psi = \le(e, \overline{e}\ri)\) is a Dirac spinor that groups together the left- and right-handed Weyl spinors \(e\) and \(\overline{e}\).
%
\(\overline{\sigma}^\mu \reppedby \le(I, \vec{\sigma}\ri)\) is a four-vector whose components are each 2\(\times\)2 matrices;
%
it is an \vocab{invariant symbol}%
\footnote{
    Invariant symbols are also known as a Clebsch-Gordan coefficients of the associated group.
    %
    See \href{https://web.physics.ucsb.edu/~mark/ms-qft-DRAFT.pdf\#page=415}{Mark Srednicki's \underline{Quantum Field Theory}, Chapters 34, 35, and 70.} \cite{Srednicki:2007qs}.
}
%
of the Lorentz group.
%
The Lagrangian of massless QED contains a free photon (gauge field), encoded by the piece \(\mathcal{L} \supset -F_{\mu\nu}F^{\mu\nu}/4\), a free fermion, encoded by \(\mathcal{L} \supset i\overline{e} \overline{\sigma}^\mu \partial_\mu e = i \bar{\Psi} \slashed{\partial} \Psi\), and an interaction \(\mathcal{L} \supset A^\mu J_\mu = i\overline{e} \overline{\sigma}^\mu A_\mu e = i \bar{\Psi} \slashed{A} \Psi\) between the photon and the electromagnetic current generated by the presence of the electron and positron.

The generalization of electromagnetism that appears in the \gls{sm} is known as \textit{Yang-Mills theory}.
%
In Yang-Mills theory, the group U(1) which encodes the gauge transformation properties of \(A_\mu\) is upgraded to a more general group (in particular, a Lie group).
%
The most common example -- and the example relevant for \gls{qcd} and the \gls{sm} -- is the group SU(N), isomorphic to the set of \(N\times N\) unitary matrices with determinant 1.

\remark{}{
    The classical properties of Yang-Mills theory, even with matter (electrons in QED, quarks in \gls{qcd}, and even the Higgs boson in the \gls{sm}), can be explained elegantly in terms of geometric concepts (connections, principal bundles, horizontal subspaces, spinor bundles, etc.).
    %
    Such an explanation is far outside the scope of our discussion, but some relevant mathematical resources include \Reffs{spinorial_chessboard,Hamilton:2017gbn}.
}

@@

The covariant derivative in the representation \(R\) also becomes a matrix, and takes the form
\begin{align}
    D_{(R)\,\mu}
    :=
    \partial_\mu
    -
    i g T^a_{(R)} A^a_\mu
    \equiv
    \partial_\mu
    -
    i g A_{(R)\,\mu}
    \,,
\end{align}
where the gauge transformation now takes the form
\begin{align}
    D^\prime_{(R)\,\mu}
    :=
    U_{(R)}
    D_{(R)\,\mu}
    U^\dagger_{(R)}
    \,,
\end{align}
and \(U_{(R)} := U_{(R)}(x)\) is dependent on the space-time position at which it is evaluated.
%
In terms of the gauge field \(A^a_\mu\), we have
\begin{align}
    T^a_{(R)} A^{\prime\,a}_\mu
    =
    A_\mu^b
    U_{(R)} T^b_{(R)} U^\dagger_{(R)}
    +
    \frac{i}{g} U_{(R)} \partial_\mu U^\dagger_{(R)}
    \,,
\end{align}
and, when the \(\alpha^a(x)\) are small enough to discard terms of \(\mathcal{O}(\alpha^2)\),
\begin{align}
    A^{\prime\,a}_\mu
    =
    A^{a}_\mu
    -
    \partial_\mu \alpha^a
    +
    g f^{abc} A^b_\mu \alpha^c
    \,.
\end{align}
% Check by explicit computation.
% also, plus sign in final term on pg 5 of Nate's 229B Lecture 12

The field strength tensor is
\begin{align}
    \label{eq:non-abelian-field-strength}
    F_{(R)\,\mu\nu}
    \equiv
    T^a_{(R)}
    F\indices{^a_{\mu\nu}}
    :=
    \frac{i}{g}
    \comm{D_{(R)\,\mu}}{D_{(R)\,\nu}}
    =
    T^a_{(R)}
    \le(
        \partial_\mu A^a_\nu - \partial_\nu A^a_\mu
        + g f^{abc} A^b_\mu A^c_\nu
    \ri)
    \,.
\end{align}
The additional non-linearity of \(F\indices{^a_{\mu\nu}}\) may appear complicated or even innocuous in \Eq{non-abelian-field-strength}, but it has incredibly deep implications for the existence of \gls{qcd} -- it is the reason \gls{qcd} overcomes the weaknesses of the quantum field theories that precede it, leads to the asymptotic freedom of quarks and gluons discussed below, and allows us to perform perturbative calculations at high energies (this is discussed further in \Sec{phd}).

Matter also matters.
%
Quarks in particular will play a large role in our discussion.
%
A \textit{matter field} \(\varphi\) -- e.g. a scalar or fermion field -- \vocab{transforms under the representation \(R\)} (of the gauge group) if
\begin{align}
    \varphi^\prime(x) = U_{(R)}(x) \, \varphi(x)
    \,.
\end{align}
%
A gauge theory with fermions \(\{\psi_{i}\}\) transforming in the representations \(R_i\) of the gauge group and scalars \(\{\phi_k\}\) transforming in the representations \(R_k\) can be described in terms of the \vocab{Yang-Mills-matter Lagrangian}%
\footnote{
    Non-Lagrangian methods for understanding \gls{qft}, such as bootstrap methods (e.g. the \textit{amplitudes program} of \gls{qft}) and the operator product expansion, are also available.
    %
    They can provide more rigorous descriptions of quantum dynamics than we pursue in this thesis, but are often more technically difficult as well.
}.
%
With massless Weyl fermions (guaranteed in the \gls{sm} due to the chiral representations of the fermionic \gls{sm} matter content), in \(d=3+1\) dimensions, and ignoring higher-dimension operators, it takes the form
\begin{subequations}
    \label{eq:yang-mills-matter-lagrangian}
\begin{align}
    \mathcal{L}
    &=
    \mathcal{L}_\text{YM}
    +
    \mathcal{L}_\text{fermion}
    +
    \mathcal{L}_\text{scalar}
    +
    \mathcal{L}_\text{Yukawa}
    \,,
    \\
    \mathcal{L}_\text{YM}
    &=
    -
    \frac{1}{2} \Tr F_{\mu\nu} F^{\mu\nu}
    -
    \frac{1}{8 \pi^2}
    \epsilon^{\mu\nu\rho\sigma}
    \Tr
        \Theta_\text{CP}
        F_{\mu\nu}
        F_{\rho\sigma}
    \,
    \\
    \mathcal{L}_\text{fermion}
    &=
    i \sum_i \psi_i^\dagger \overline{\sigma}^\mu D_{(R_i)\,\mu} \psi_i
    \\
    \mathcal{L}_\text{scalar}
    &=
    \sum_k \le(D_{(R_k)}^\mu \phi_k\ri)^\dagger D_{(R_k)\,\,\mu} \phi_k
    -
    \sum_{k,\ell} \! {}^\prime
    \,\,
    m_{k\ell}^2
    \,\,
    \phi_k \, \phi_\ell
    \,\,
    S_{R_k R_\ell}
    +
    \text{h.c}
    \\
    \notag
    &\qquad\qquad
    +
    \sum_{k,\ell,m,n} \!\!\!\! {}^\prime
    \,\,
    \lambda_{k\ell m n}
    \,\,
    \phi_k \, \phi_\ell \, \phi_m \, \phi_n
    \,\,
    S_{R_k R_\ell R_m R_n}
    +
    \text{h.c.}
    \\
    \mathcal{L}_\text{Yukawa}
    &=
    \sum_{ij}
    \sum_{k} \! {}^\prime
    \,\,
    y_{ijk}
    \,\,
    \psi_i \,\psi_j \,\, \phi_k
    \,\,
    S_{R_i R_j R_k}
    +
    \text{h.c.}
    \,,
\end{align}
\end{subequations}
where the \(\sum'\) indicates a sum on both the scalar fields \(\phi_k\) and the complex conjugates \(\phi_k^\dagger\), the \(S\) are invariant symbols of the gauge group which project onto singlet representations and guarantee the invariance of the Lagrangian under gauge transformations, the CP-violating angle \(\Theta_\text{CP}\) -- usually multiplied by an additional factor of \(g^2\) -- is a (hermitian) central charge of the gauge group (represented in the adjoint representation), and ``h.c.'' means hermitian conjugate.

\remark{}{
    A single degree of freedom of the central charge \(\Theta_\text{CP}\) can be removed for every anomalous global symmetry of \(\mathcal{L}\), and CP-violating angles associated with Abelian (U(1) or \(\mathbb{R}\)) factors of the gauge group can be ignored in topologically trivial spaces.
    %
    See, for example, \Reff{Tong:2017oea}.
}

@@

\remark{}{
    There are additional non-trivial constraints on the matter content of a \gls{qft}, called \emph{gauge anomaly cancellation conditions}, which require that the charges of the matter fields are delicately balanced.
    %
    We will consider \gls{qcd} without the top quark in this thesis, and, though this effective theory \textit{does} have an anomaly, it is nonetheless consistent.
    %
    For an understanding of the properties of effective field theories which appear to carry anomalies, see the incredibly lucid account of \Reff{Preskill:1990fr}.
}
